Report on the Third Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE3)

### 2015-10-28 WSSSPE3/report/wg_appendix_software_credit.tex
small fixes/comments to software credit appendix

At a minimum, the metadata needed for software citation (aside from the DOI for the software itself) includes:
\begin{itemize}
    \item Contributors, in the form of name and ORCID,
    \item Software dependencies, in the form of DOI, and
    \item Other people and artifacts that would be cited or acknowledged in a paper.
\end{itemize}
This information would then be contained in a citation file as part of 
the, e.g., GitHub repository.
The group also discussed similar efforts such as Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an attempt to codify minimal 
metadata schemes in JSON and XML for scientific software and code, and 
implementing transitive credit via JSON-LD~\cite{wssspe2_katz}.
Some questions arose about how this information would be stored for 
closed-source software.

@@

Although far from a standard practice, examples of citing software in 
publications can be found in various scientific communities---notably, 
representative samples can be found in astronomy\footnote{Astronomy software citation examples: \url{https://docs.google.com/document/d/1q9ULl7alA3veL7Qwg7jGteRWeJwlrkvRHSXjvt-rTs0}} and biology~\cite{Howison2015}.
The group recommended collecting similar examples from other communities, then 
developing a software citation principles document in concert with the 
FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details) 
following the model of the FORCE11 Data Citation Principles 
document~\cite{DataCitation2014}.

The group also discussioned briefly whether software used directly in 
a publication---whether to perform simulation, analysis, or as a dependency 
for new software developed---should be distinguished from other references 
due to the dependence of the study on these research artifacts.
Suggestions included a separate list of citations (with DOIs) for software 
and other research objects that serve this sort of ``vital'' role.
Similar recommendations were made by the credit breakout group at 
WSSSPE2~\cite{WSSSPE2}.

Finally, although a discrete task from the software citation, significant 
discussion focused on ensuring software citations are indexed in the same 
manner as publications---allowing the construction of a corresponding 
software citation network.
Currently, software releases can receive citable DOIs via 
Zenodo\footnote{Zenodo, \url{https://zenodo.org}} and 
FigShare\footnote{FigShare: \url{https://figshare.com}}; however, these 
citations are not processed by indexers such as Web of Science, Scopus, 
or Google Scholar.
Thus, either in parallel or following the primary task, the group will 
need to reach out to these organizations 
Initial conversations with Elsevier\slash Scopus (via Michael Taylor) 
during WSSSPE3 clarified that Scopus is not yet DataCite DOI aware, and 
also does not yet have an internal identifier for software or data (but 
needs\slash plans to add this support).
According to Taylor, they prefer a ``software article'' with the usual 
article metadata (e.g., authors, citations), and mentioned Zenodo as an 
example---this proposal seemed to align with group discussions.
Taylor also mentioned another benefit of the software and associated DOI 
on GitHub: in addition to a citation, could obtain statistics on 
usage/downloads/forks.

@@

Modern scientific research relies considerably on software, and ensuring 
that developers of such scientific software receive credit for their efforts 
will encourage additional creation and maintenance.
There currently is no standard mechanism for citing software or receiving credit for software (akin to publications/citations).
Software is eligible for DOI assignment, but DOI metadata fields are not well tuned or standardized for software (vs. publications).
Some software providers apply for DOIs but it is still not widely adopted.
Also, there is no mechanism to cite software dependencies within software.

Major obstacles include the fact that indexers (e.g., Scopus, Web of Science, 
Google Scholar) do not currently support software/data document types or 
DataCite DOIs.
Therefore, even with univeral association of scientific software with DOIs 
and standardizing practices for citing software in publications

Although the working group discussions at WSSSPE3 did not focus much on the 
topic of tenure, the group recognized that there is no standard policy---even 
within a single university---for software products to be included in 
promotion and tenure dossiers.
Thus, it may be difficult to encourage valuing software contributions across 
the United States or United Kingdom, and globally; furthermore, stakeholders 
are typically not tenured and thus may not be influential enough.

>>>>

At a minimum, the metadata needed for software citation (aside from the DOI for the software itself) includes:
\begin{itemize}
    \item Contributors, in the form of name and ORCID,
    \item Software dependencies, in the form of DOI, and
    \item Other people and artifacts that would be cited or acknowledged in a paper.
\end{itemize}
\katznote{is a name/title also needed?}
This information would then be contained in a citation file, e.g.,  as part of 
the GitHub repository.
The group also discussed similar efforts such as Codemeta\footnote{Codemeta: \url{https://github.com/codemeta/codemeta}}, an attempt to codify minimal 
metadata schemes in JSON and XML for scientific software and code, and 
implementing transitive credit via JSON-LD~\cite{wssspe2_katz}.
Some questions arose about how this information would be stored for 
closed-source software.

@@

Although far from a standard practice, examples of citing software in 
publications can be found in various scientific communities---notably, 
representative samples can be found in astronomy\footnote{Astronomy software citation examples: \url{https://docs.google.com/document/d/1q9ULl7alA3veL7Qwg7jGteRWeJwlrkvRHSXjvt-rTs0}} and biology~\cite{Howison2015}.
The group recommended collecting similar examples from other communities, then 
developing a software citation principles document in concert with the 
FORCE11 Software Citation Working Group (see \S\ref{SC:plan} for more details), 
following the model of the FORCE11 Data Citation Principles 
document~\cite{DataCitation2014}.

The group also discussed briefly whether software used directly in 
a publication---whether to perform simulation, analysis, or as a dependency 
for new software developed---should be distinguished from other references 
due to the dependence of the study on these research artifacts.
Suggestions included a separate list of citations (with DOIs) for software 
and other research objects that serve this sort of ``vital'' role.
Similar recommendations were made by the credit breakout group at 
WSSSPE2~\cite{WSSSPE2}.

Finally, although a discrete task from software citations themselves, significant 
discussion focused on ensuring software citations are indexed in the same 
manner as publications, allowing the construction of a corresponding 
software citation network.
Currently, software releases can receive citable DOIs via 
Zenodo\footnote{Zenodo, \url{https://zenodo.org}} and 
figshare\footnote{figshare: \url{https://figshare.com}}; however, these 
citations are not processed by indexers such as Web of Science, Scopus, 
or Google Scholar.
Thus, either in parallel or following the primary task, the group will 
need to reach out to these organizations 
Initial conversations with Elsevier\slash Scopus (via Michael Taylor) 
during WSSSPE3 clarified that Scopus is not yet DataCite DOI aware, and 
also does not yet have an internal identifier for software or data (but 
needs\slash plans to add this support).
Taylor said they prefer a ``software article'' with the usual 
article metadata (e.g., authors, citations), and mentioned Zenodo as an 
example -- this proposal seemed to align with group discussions.
Taylor also mentioned another benefit of the software and associated DOI 
on GitHub: in addition to a citation, could obtain statistics on 
usage/downloads/forks, which happens to be what Depsy\footnote{Depsy: \url{https://depsy.org}} is beginning to try to do.

@@

Modern scientific research relies considerably on software, and ensuring 
that developers of such scientific software receive credit for their efforts 
will encourage additional creation and maintenance. \katznote{the prev sentence is good, but feels a bit odd in the middle of the appendix, rather than at the beginning}
There currently is no standard mechanism for citing software or receiving credit for software (akin to publications/citations).
Software is eligible for DOI assignment, but DOI metadata fields are not well tuned or standardized for software (vs. publications).
Some software providers apply for DOIs but it is still not widely adopted.
Also, there is no mechanism to cite software dependencies within software.

Major obstacles include the fact that indexers (e.g., Scopus, Web of Science, 
Google Scholar) do not currently support software/data document types or 
DataCite DOIs.
Therefore, even with universal association of scientific software with DOIs 
and standardizing practices for citing software in publications \katznote{this sentence is missing its ed}

Although this working group's discussions at WSSSPE3 did not focus much on the 
topic of tenure, the group recognized that there is no standard policy---even 
within a single university---for software products to be included in 
promotion and tenure dossiers.
Thus, it may be difficult to encourage valuing software contributions across 
the United States or United Kingdom and globally; furthermore, stakeholders 
are typically not tenured and thus may not be influential enough. \katznote{to do what?}
